
\begin{abstract}

  Distributed real-time editors made real-time editing easy for
  millions of users. However, main stream editors relies on cloud
  services to mediate sessions raising privacy and scalability
  issues. Decentralized editors tackles privacy issues but scalability
  issues remains. We aim to build a decentralized editor that allow
  real-time editing anytime, anywhere, whatever the number of
  participants. In this paper, we propose an approach based on a
  massively replicated sequence structure that represent the shared
  document. We establish an original tradeoff on communication, time
  and space complexity to maintain this sequence over a network of
  browsers. We prove a sublinear upper-bound on communication
  complexity while preserving an affordable time and space
  complexity. In order to validate this trade-off, we built a full
  working editor and measure its performance on a large scale
  experiments involving up till 600 participants. The results show a
  traffic increasing as $\mathcal{O}((\log I)^2\ln R)$ where $I$ is
  the number of insertions in the document, and $R$ the number of
  participants. 


---8<----

  Distributed real-time editors such as Google Docs, or Etherpad allow users
  distributed in space and organizations to collaborate easily with only web
  browsers. Yet, main stream editors rely on central servers with issues in
  privacy, single point of failure and scalability.  Decentralized editors
  settle privacy problems but scalability issues remain.  None of existing
  algorithms provide a satisfying complexity in both communication and space.
  In this paper, we provide the complexity analysis of \LSEQ, an algorithm
  generating identifiers enjoying a polylogarithmic upper bound on space
  complexity which in turns directly impacts traffic. We provide the outline to
  build a decentralized real-time editor. We build an editor following these
  guidelines and evaluate it on large scale experiments involving up till 600
  participants. The results show a traffic increasing as
  $\mathcal{O}((\log I)^2\ln R)$ where $I$ is the number of insertions in the
  document, and $R$ the number of participants. As such, the editor scales well
  in both these dimensions and does not require running consensus protocols
  known to be overly expensive in distributed systems.

  % In this paper, we propose \CRATE; a working decentralized real-time editor
  % that supports large number of users working on large documents. In particular,
  % \CRATE achieves a sublinearly upper-bounded communication complexity in
  % $\mathcal{O}((\log d)^2\ln{m})$ where $d$ is the document size and $m$ the
  % membership size. Thanks to WebRTC, \CRATE can be easily deployed on a network
  % of browsers.  We evaluate \CRATE on large-scale experiments in the Grid'5000
  % testbed involving uptill 600 participants. As expected, we observe a
  % logarithmic progression of the generated traffic according to the size of the
  % shared document and the number of participants.
\end{abstract}

%%% Local Variables:
%%% mode: latex
%%% TeX-master: "../paper"
%%% End:
